Factors affecting reliablity of state-space age-structured assessment models

### 2024-01-25 Project_0/paper/paper.Rmd
draft CJFAS abstract 175 words

@@

Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}. State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models \citep{hoyleetal22, punt23}.

>>>>

## Abstract {-}

<!-- note CJFAS abstract word limit 175, so I have drafted a shorter summary here -->

Random effects can be included in state-space assessment models on many processes and in many ways, and guidance is needed on statistical reliability and model selection criteria. We simulated 72 operating models with varying fishing history, observation error uncertainty, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality). We fit estimating models with different assumptions on the process error source, whether (mean) natural mortality was estimated, and whether a stock-recruit relationship was estimated.

Models that assumed the correct process error source had high convergence and low bias. Bias was also low under most process error assumptions when there was contrast in fishing pressure. Stock-recruitment parameters were only reliably estimated in ideal situations. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types. Retrospective patterns were generally weak except for recruitment when observation error was high, even with the correct process error assumptions. When models did exhibit some retrospective pattern, estimating natural mortality often removed it.

@@

Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US \citep{nielsenberg14,cadigan16,pedersenberg17,stockmiller21}. State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models \citep{hoyleetal22, punt23}.
